---
title: "Introduction to Machine Learning"
format: 
  html:
    embed-resources: true
---

## Intro to the datasets

Property sales information as described in De Cock (2011).

* problem type: supervised regression
* response variable: Sale_Price (i.e., $195,000, $215,000)
* features: 80
* observations: 2,930
* objective: use property attributes to predict the sale price of a home
* access: provided by the `AmesHousing` package (Kuhn 2017a)
* more details: See `?AmesHousing::ames_raw`

```{r}
#| output: false

### access data
ames <- AmesHousing::make_ames()

### initial dimension
dim(ames)
# [1] 2930   81

### response variable
head(ames$Sale_Price)
# [1] 215000 105000 172000 244000 189900 195500
```

Employee attrition information originally provided by IBM Watson Analytics Lab.

* problem type: supervised binomial classification
* response variable: Attrition (i.e., “Yes”, “No”)
* features: 30
* observations: 1,470
* objective: use employee attributes to predict if they will attrit (leave the company)
* access: provided by the `rsample` package (Kuhn and Wickham 2019)
    * NOTE: these are now in the `modeldata` package!
* more details: See `?rsample::attrition`
    * see `?modeldata::attrition` instead

```{r}
#| output: false

### access data
attrition <- modeldata::attrition

### initial dimension
dim(attrition)
# [1] 1470   31

### response variable
head(attrition$Attrition)
# [1] Yes No  Yes No  No  No 
# Levels: No Yes
```

Image information for handwritten numbers originally presented to AT&T Bell Lab’s to help build automatic mail-sorting machines for the USPS. Has been used since early 1990s to compare machine learning performance on pattern recognition (i.e., LeCun et al. (1990); LeCun et al. (1998); Cireşan, Meier, and Schmidhuber (2012)).

* Problem type: supervised multinomial classification
* response variable: V785 (i.e., numbers to predict: 0, 1, …, 9)
* features: 784
* observations: 60,000 (train) / 10,000 (test)
* objective: use attributes about the “darkness” of each of the 784 pixels in images of handwritten numbers to predict if the number is 0, 1, …, or 9.
* access: provided by the `dslabs` package (Irizarry 2018)
* more details: See `?dslabs::read_mnist()` and online MNIST documentation


```{r}
#| output: false

### access data
mnist <- dslabs::read_mnist()
names(mnist)
# [1] "train" "test"

### initial feature dimensions
dim(mnist$train$images)
# [1] 60000   784

### response variable
head(mnist$train$labels)
# [1] 5 0 4 1 9 2
```


Grocery items and quantities purchased. Each observation represents a single basket of goods that were purchased together.

* Problem type: unsupervised basket analysis
* response variable: NA
* features: 42
* observations: 2,000
* objective: use attributes of each basket to identify common groupings of items purchased together.
* access: available on the companion website for this book

```{r}
#| output: false

### URL to download/read in the data
url <- "https://koalaverse.github.io/homlr/data/my_basket.csv"

### Access data
my_basket <- readr::read_csv(url)

### Print dimensions
dim(my_basket)
# [1] 2000   42

### Peek at response variable
my_basket
# # A tibble: 2,000 x 42
#    `7up` lasagna pepsi   yop red.wine cheese   bbq bulmers mayonnaise
#    <dbl>   <dbl> <dbl> <dbl>    <dbl>  <dbl> <dbl>   <dbl>      <dbl>
#  1     0       0     0     0        0      0     0       0          0
#  2     0       0     0     0        0      0     0       0          0
#  3     0       0     0     0        0      0     0       0          0
#  4     0       0     0     2        1      0     0       0          0
#  5     0       0     0     0        0      0     0       2          0
#  6     0       0     0     0        0      0     0       0          0
#  7     1       1     0     0        0      0     1       0          0
#  8     0       0     0     0        0      0     0       0          0
#  9     0       1     0     0        0      0     0       0          0
# 10     0       0     0     0        0      0     0       0          0
# # … with 1,990 more rows, and 33 more variables: horlics <dbl>,
# #   chicken.tikka <dbl>, milk <dbl>, mars <dbl>, coke <dbl>,
# #   lottery <dbl>, bread <dbl>, pizza <dbl>, sunny.delight <dbl>,
# #   ham <dbl>, lettuce <dbl>, kronenbourg <dbl>, leeks <dbl>, fanta <dbl>,
# #   tea <dbl>, whiskey <dbl>, peas <dbl>, newspaper <dbl>, muesli <dbl>,
# #   white.wine <dbl>, carrots <dbl>, spinach <dbl>, pate <dbl>,
# #   instant.coffee <dbl>, twix <dbl>, potatoes <dbl>, fosters <dbl>,
# #   soup <dbl>, toad.in.hole <dbl>, coco.pops <dbl>, kitkat <dbl>,
# #   broccoli <dbl>, cigarettes <dbl>
```
